---
title: "Mapping with R"
author: 
- "Kent Johnson"
- "kent@kentsjohnson.com"
date: '`r Sys.Date()`'
output: 
  slidy_presentation:
    highlight: pygments
    css: styles.css
---

```{r parameters, echo=FALSE,include=FALSE,message=FALSE}
knitr::opts_chunk$set(echo=TRUE,fig.width=8, fig.height=4.5, comment=NA, warning=FALSE, message=FALSE)
options(width=100)
```

## Maps in R?

> - __Good__
> - R has good support for spatial data including shapefiles, GeoJSON, spatial operations
> - Support for data manipulation, general graphics, statistics
> - Great reporting capabilities with `knitr` and `rmarkdown`  
  This presentation was authored in R and markdown
> - Capable programming language
> - Use `knitr` and `leaflet` to embed interactive maps in HTML pages
> - Growing support for interactive HTML reports with no JavaScript coding required!

> - __Not so good__
> - It's programming :-)
> - R is a quirky programming language and can take some time to learn
> - Interactive capabilities are less than what you can do directly in Javascript

## Today's goal

Make a map showing Cambridge properties built since 1980, color-coded by year built.

Data is from the Cambridge Open Data portal (http://data.cambridgema.gov).

> - Assessor's data gives parcel address, year built
> - Parcel shapefile contains the parcel outlines

## Libraries

```{r}
library(rgdal)
library(leaflet)
```

> - `rgdal` package supports reading and writing shapefiles
> - `leaflet` package (http://rstudio.github.io/leaflet/) connects to `leaflet.js`
> - Install with `install.packages(c('rgdal', 'leaflet'))` or the _Packages_ tab in RStudio

## Load the assessor's data

```{r}
# https://data.cambridgema.gov/Assessing/Assessing-Building-Information/crnm-mw9n
ass2015 = read.csv('data/Assessing_Building_Information.csv', 
               stringsAsFactors=FALSE)
```

> - `read.csv` loads a .csv file into an R data.frame.
> - `stringsAsFactors=FALSE` prevents conversion of string columns into categorical
variables.

## Explore assessor's data

```{r}
nrow(ass2015)
names(ass2015)
```
> - `nrow` shows the number of rows in the table
> - `names` shows the column names in the table
> - Notice all special characters in column names have been changed to periods.

## Explore assessor's data

```{r}
summary(ass2015$Actual.Year.Built)
head(table(ass2015$Actual.Year.Built), n=50)
# table(ass2015$Actual.Year.Built)
```
> - `summary` gives a simple summary of numerical data
> - `table` tabulates values in the year built column.
> - In RStudio: `View(ass2015)` or clicking the variable in _Environment_ gives a sortable, searchable table view

## Filter out old & duplicate properties

```{r}
ass_new = subset(ass2015, Actual.Year.Built>=1980) 
ass_new = ass_new[!duplicated(ass_new$GIS.ID),]
```

> - `subset` selects properties built in 1980 or later
> - `duplicated` finds entries which have already appeared in a vector
> - There may be multiple properties on one parcel (e.g. condos)
> - We just need one property to identify the parcel

## Explore
```{r fig.width=7, fig.height=3.5}
table(ass_new$Actual.Year.Built)
hist(ass_new$Actual.Year.Built)
```

## Read parcel shapefile

```{r}
# https://data.cambridgema.gov/Assessing/Parcels-Fiscal-Year-2015/rst6-227j
parcels2015 = readOGR('data/ASSESSING_ParcelsFY2015.shp', 
        'ASSESSING_ParcelsFY2015', stringsAsFactors=FALSE)

names(parcels2015)
# View(parcels2015@data)
```

> - `readOGR` returns a `SpatialPolygonsDataFrame` which combines the parcel shapes with a data.frame
> - `parcels2015@data` is the data.frame
> - `parcels2015@polygons` contains the polygon data.

## Merge

```{r}
new_parcels = parcels2015[parcels2015$ML %in% ass_new$GIS.ID,]
new_parcels = merge(new_parcels, ass_new, by.x='ML', by.y='GIS.ID')
```

> - Subset the parcel data to just the ones we want
> - Parcel ID is `ML` (map-lot) in the parcel data and `GIS.ID` in the assessor's data.
> - Use indexing to select (could also use `subset`)
> - `%in%` creates a boolean vector
> - Merge property data with parcel shapes to make a single data set

## A simple map

The `plot` method will make a simple map of spatial data.

```{r}
plot(new_parcels)
```

## A simple map
> - It's not much but it verifies that we have reasonable spatial data
> - We could add color, political boundaries, etc. to make a real map
> - Slippy maps are more fun!

## Transform to WGS 84 for web mapping

```{r transform}
new_osm = spTransform(new_parcels, CRS("+init=EPSG:4326"))
```

> - Cambridge data is in Massachusetts State Plane coordinates
> - Web maps use WGS 84 _aka_ epsg:4326
> - If your data doesn't show up, this might be the reason

## First slippy map

The R API to leaflet.js is patterned on the Javascript API.

```{r first_map, results='hide'}
leaflet() %>% 
  addTiles() %>% 
  setView(-71.128184, 42.3769824, zoom = 14) %>% 
  addPolygons(data=new_osm)
```

> - `leaflet()` creates the map object
> - `addTiles()` creates the background tiled layer
> - `setView()` sets the initial location and zoom
> - `addPolygons()` adds our data to the map as a polygon overlay
> - ` %>%` is a pipe operator

## First slippy map

```{r ref.label='first_map', echo=FALSE, results='markup'}
```

## Change the base layer

Stamen Design's `toner-lite` is a good base map, it stays out of the way.

```{r}
base_map = leaflet() %>% 
  addTiles('http://{s}.tile.stamen.com/toner-lite/{z}/{x}/{y}.jpg',
    attribution = 
      paste('Map tiles by <a href="http://stamen.com">Stamen Design</a>,',
      '<a href="http://creativecommons.org/licenses/by/3.0">CC BY 3.0</a> &mdash;',
      'Map data &copy; <a href="http://openstreetmap.org">OpenStreetMap</a>',
      'contributors, <a href="http://creativecommons.org/licenses/by-sa/2.0/">',
      'CC-BY-SA</a>')) %>% 
  setView(-71.128184, 42.3769824, zoom = 14) 
```

> - Save the base map in a variable to reduce repetition later
> - `addTiles()` takes a URL template
> - __Many__ tile providers, see http://leaflet-extras.github.io/leaflet-providers/preview/
> - Include proper attribution!

## Change the base layer

```{r}
base_map %>% 
  addPolygons(data=new_osm)
```

## Color

```{r}
palette = c('#fed976', '#feb24c', '#fd8d3c', 
            '#fc4e2a', '#e31a1c', '#bd0026', '#800026')

year_built = new_parcels@data$Actual.Year.Built
colors = palette[cut(year_built, c(seq(1980, 2010, 5), 2015), 
                     include.lowest=TRUE, labels=FALSE)]
```

> - `palette` is the actual colors  
    From the 9-class YlOrRd ColorBrewer palette  
    http://colorbrewer2.org/?type=sequential&scheme=YlOrRd&n=9
> - `colors` is a vector of colors, one per property, selected from `palette`
> - Use one color per five-year period
> - `cut` divides the years into ranges

## Color

```{r}
base_map %>% 
  addPolygons(data=new_osm, 
              color='black', fillColor=colors) # Colors
```

## Popups

- Popups are HTML, build them in R using `apply` and `paste`

```{r}
# Make popups showing the location and year built.
popups = apply(new_parcels@data, 1, function(parcel) {
         paste0('<p>', parcel['Location'], '<br>',
         'Year: ', parcel['Actual.Year.Built'], '<br></p>')
         })
```

> - `apply` performs an operation on each element along one dimension (here, for each row)
> - `paste0` concatenates strings

## Popups

```{r}
base_map %>% 
  addPolygons(data=new_osm, 
              color='black', fillColor=colors, # Colors
              popup=unname(popups),            # Popups
              options=c(pathOptions(), popupOptions(minWidth=100)))
```

## A little more styling and we're done

See the `leaflet.js` docs for available options.

```{r final, results='hide'}
base_map %>% 
  addPolygons(data=new_osm, 
              color='black', fillColor=colors, # Colors
              weight=1, dashArray=1,           # Line weight
              opacity=0.7, fillOpacity=0.5,    # Transparency
              popup=unname(popups),            # Popups
              options=c(pathOptions(), popupOptions(minWidth=100)))
```

## Final map

```{r ref.label='final', echo=FALSE, results='markup'}
```

## Links

- R home page  
  https://www.r-project.org/
- RStudio IDE  
  https://www.rstudio.com/products/RStudio/
- `leaflet` package for R  
  http://rstudio.github.io/leaflet/
- A more detailed tutorial on spatial analysis and mapping in R  
  http://robinlovelace.net/r/2014/01/30/spatial-data-with-R-tutorial.html
- More on static maps  
  http://www.molecularecologist.com/2012/09/making-maps-with-r/

### Questions?